Typed Lua: An Optional Type System for Lua

### 2014-03-10 doc/paper/typedlua.tex
Updating expected contributions

\newcommand{\Any}{\mathbf{any}}
\newcommand{\Top}{\mathbf{value}}
\newcommand{\Nil}{\mathbf{nil}}
\newcommand{\False}{\mathbf{false}}
\newcommand{\True}{\mathbf{true}}
\newcommand{\Boolean}{\mathbf{boolean}}
\newcommand{\Number}{\mathbf{number}}
\newcommand{\String}{\mathbf{string}}
\newcommand{\Void}{\mathbf{void}}

@@

Dynamically typed languages such as Lua avoid static types in favor of
run-time {\em type tags} that classify the values they compute, and
their implementations use these tags to perform run-time (or dynamic)
checking and guarantee that only valid operations are
performed~\citep{pierce2002tpl}.

The absence of static types means that programmers do not need to
bother about abstracting types that might require a complex type
system and type checker to validate, leading to simpler and more flexible
languages and implementations. But this absence may also hide bugs
that will be caught only after deployment if programmers do not properly
test their code.

In contrast, static type checking helps programmers detect many 
bugs during the development phase. Static types also provide a
conceptual framework that helps programmers define modules
and interfaces that can be combined to structure the development
of large programs.

The early error detection and better program structure afforded by
static type checking can lead programmers to migrate their code from
a dynamically typed to a statically typed language, once their simple
scripts become complex programs~\citep{tobin-hochstadt2006ims}.
As this migration involves languages with different syntax and
semantics, it requires a complete rewrite of existing programs intead
of incremental evolution from dynamic to static types.

Ideally, programming languages should offer programmers the
option to choose between static and dynamic typing:
\textit{optional type systems}~\citep{bracha2004pluggable} and
\textit{gradual typing}~\citep{siek2006gradual} are two approaches
that offer programmers the option to use type annotations where static
typing is needed, incrementally migrating a system from dynamic
to static types. The difference between these two approaches is the
way they treat run-time semantics: while optional type systems
do not affect the run-time semantics,
gradual typing uses run-time checks to ensure that dynamically typed
code does not violate the invariants of statically typed code.

In this work we outline the design of Typed Lua:
an optional type system for Lua that is complex enough to
preserve some of the idioms that Lua programmers are already used to,
while adding new constructs that help programmers structure Lua
programs.

Lua is a small imperative scripting language that prefers to provide
mechanisms instead of fixed policies for structuring programs,
due to its history as an embedded extension language. Even features
such as a module system and object orientation are a matter of 
convention instead of built-in features.

The lack of standard policies
is a challenge for the design of a static type system for the Lua
language, and our approach was to extract the most common
idioms from a large corpus of Lua libraries as a guide to what
Typed Lua should focus on, and setting fixed policies for organizing
a program in modules and using class-based object orientation.

Typed Lua allows statically typed Lua code to coexist and interact with 
dynamically typed code. The Typed Lua compiler warns the programmer
about type errors, but always generates Lua code that runs in unmodified
Lua implementations. The programmer can enjoy some of the benefits
of static types even without converting existing Lua modules to typed Lua:
a dynamically typed module can export a statically typed interface, and
statically typed users of the module will have their use of the module
checked by the compiler.

Unlike gradual type systems, Typed Lua does not insert run-time
checks between dynamically and statically typed parts of the program.
Unlike some optional type systems, the statically typed subset of
Typed Lua is sound by design, so a later version of the Typed Lua
compiler can swicth to gradual instead of just optional typing.

In Section~\ref{sec:proposal} we outline the key parts in
the design of Typed Lua. In Section~\ref{sec:review} we review
related work on mixing static and dynamic typing in the same language.
Finally, Section~\ref{sec:con} gives some concluding remarks and
our plans for the future of Typed Lua.

Lua is an imperative scripting language with first-class functions
(with proper lexical scoping) where the main data structure is the
{\em table}, an associative array that can play the part of arrays,
records, maps, objects, etc., with syntactic sugar and
metaprogramming through operator overloading built into
the language. Unlike other scripting languages, Lua has very
limited coercion among different data types.

The primary use of Lua has always been as an embedded language
for configuration and extension of other applications, and the result
is a fragmented ecosystem of libraries, and different ideas among
Lua programmers on how they should use the language features and
how they should structure programs.

The design of Typed Lua is informed by a (mostly automated) survey
of Lua idioms used in a large corpus of Lua libraries, instead of
relying just on the semantics of the language. We cover the main parts
of the design, along with how they relate to Lua, in the rest of this section,
and discuss some of what was left out in the current version of the design
in the conclusion of the paper.

\subsection{Atomic Types and Functions}

Lua values can have one of eight {\em tags}: {\em nil}, {\em boolean},
{\em number}, {\em string}, {\em function}, {\em table}, {\em userdata},
and {\em thread}. We will see how Typed Lua assigns types to
values of the first five in this section.

Figure~\ref{fig:typelang} gives the abstract syntax of Typed Lua
types. Only {\em first-class types} correspond to actual Lua
values; {\em second-class types} correspond to expression lists,
and Typed Lua uses them to type multiple assignment and function
application. Types are ordered by a subtype relationship, where
any first-class type is a subtype of $\Top$.

The {\em dynamic type} $\Any$ allows dynamically typed code to
interoperate with statically typed code; it is a subtype of $\Top$, but
neither a supertype nor a subtype of any other type. We relate $\Any$
to other types with the {\em consistency} and {\em consistent-subtype}
relationships used by gradual type systems. In practice, we can
pass a value of the dynamic type anytime we want a value of some
other type, and can pass any value where a value of the dynamic type
is expected, but these operations are tracked by the type system,
and the programmer can choose to be warned about them.

\begin{figure}[!ht]
\textbf{Type Language}\\
\dstart
$$
\begin{array}{rlr}
\multicolumn{3}{c}{\textbf{First-class types}} \\
T ::= & \;\; L & \textit{literal types}\\
& | \; B & \textit{base types}\\
& | \; \Top & \textit{top type}\\
& | \; \Any & \textit{dynamic type}\\
& | \; T \cup T & \textit{disjoint union types}\\
& | \; S \rightarrow S & \textit{function types}\\
& | \; \{T:T, ..., T:T\} & \textit{table types}\\
& | \; X & \textit{type variables}\\
& | \; \mu X.T & \textit{recursive types}\\
& | \; X_i & \textit{projection types}\\
L ::= & \multicolumn{2}{l}{\False \; | \; \True \; | \; {<}{\it number}{>} \; | \; {<}{\it string}{>} } \\
B ::= & \multicolumn{2}{l}{\Nil \; | \; \Boolean \; | \; \Number \; | \; \String} \\
\multicolumn{3}{c}{} \\
\multicolumn{3}{c}{\textbf{Second-class types}} \\
S ::= &  \multicolumn{2}{l}{\Void \; |\; V \; | \; S \cup S} \\
V ::= & \;\; T & \\
& | \; T* & \textit{vararg types}\\
& | \; T \times V & \textit{tuple types}
\end{array}
$$
\dend
\caption{Abstract syntax of Typed Lua types}
\label{fig:typelang}
\end{figure}

Typed Lua allows optional type annotations in variable function
declarations. It assigns the dynamic type to any parameter that
does not have an type annotation, but assigns more precise
types to unnanotated variables, based on the type of the
expression that gives the initial value of the variable.

In the following example, we use type annotations in a function
declaration but do not use type annotations in the declaration of a
local variable:
\begin{verbatim}
    local function factorial(n: number): number
      if n == 0 then
        return 1
      else
        return n * factorial(n - 1)
      end
    end
    local x = 5
    print(factorial(x))
\end{verbatim}

The compiler assigns the type $\Number$ to the local variable \texttt{x},
and this example compiles without warnings. Typed Lua allows
programmers to combine annotated code with
unannotated code, as we show in the following example:
\begin{verbatim}
    local function abs(n: number)
      if n < 0 then
        return -n
      else
        return n
      end
    end

    local function distance(x, y)
      return abs(x - y)
    end
\end{verbatim}

The compiler assigns the dynamic type $\Any$ to the input
parameters of \texttt{distance} because they do not have type annotations.
Subtracting a value of type $\Any$ from another also yields a value of
type $\Any$ (Lua has operator overloading, so the minus operation
is not guaranteed to return a number in this case), but
consistent subtyping lets us pass a value of type $\Any$ to a
function that expects a $\Number$.

Even though the return types
of both {\tt abs} and {\tt distance} are not given, the compiler is able to
infer a return type of $\Number$ to both functions, as they are
local and not recursive.

Lua has first-class functions, but they have some peculiarities. First,
the number of arguments passed to a function does not need to
match the function's arity; Lua silently drops extra arguments after
evaluating them, or passes {\tt nil} in place of any missing arguments.
Second, functions can return any number of values, and the number
of values returned may not be statically known. Third, Lua also has
multiple assignment, and the semantics of argument passing match
those of multiple assignment (or vice-versa); calling a function is like
doing a multiple assigment where the left side is the parameter list
and the right side is the argument list.

Typed Lua uses {\em second-class types} to encode the peculiarities
of argument passing, multiple returns, and multiple assignment. We call
them second-class because these types do not correspond to actual
values and cannot be assigned to variables or parameters: they are an
artifact of the interaction between the type system and the semantics of Lua.

As we saw in Figure \ref{fig:typelang}, a second-class type in
Typed Lua can be the type $\Void$ (the empty tuple),
a tuple of first-class types optionally ending in a variadic type,
or a union of these tuples. A variadic type $T*$ is a generator for a
sequence of values of type $T \cup \Nil$. Unions of tuples play
an important part in functions that are overloaded on the return type,
together with {\em projection types}. Both are explained in the next
section.

Typed Lua always adds a
variadic tail to the parts of a function type if none is specified, to match
the semantics of Lua function calls. In the examples above, the types
of {\tt factorial} and {\tt abs} are actually $\Number \times \Top *
 \rightarrow \Number \times \Nil *$, and the type of {\tt distance} is
$\Any \times \Any \times \Top * \rightarrow \Number \times \Nil *$.

If we call {\tt abs} with extra arguments, Typed Lua silently ignores
them, as the type signature lets {\tt abs} receive any number of extra
arguments. If we call {\tt abs} in the right side of an assignment to
more than one lvalue, Typed Lua checks if the first lvalue has a type
consistent with $\Number$, and any other lvalues need to have a type
consistent with $\Nil$.

A variadic type can only appear in the tail position of a tuple,
because Lua takes only the first value of any expression that appears
in an expression list that is not in tail position. The following example
shows the interaction between multiple returns and expression lists:

\begin{verbatim}
    local function multiple()
      return 2, "foo"
    end

    local function sum(x: number, y: number)
      return x + y
    end

    local x, y, z = multiple(), multiple()
    print(sum(multiple(), multiple())
\end{verbatim}

Function {\tt multiple} is
$\Top * \rightarrow \Number \times \String \times \Nil *$,
and {\tt sum} is $\Number \times \Number \times \Top * \rightarrow
\Number \times \Nil *$. In the right side of the multiple assignment,
only the first value produced by the first call to multiple gets used, so
the type of the right side is $\Number \times \Number \times \String \times \Nil*$,
and the types assigned to {\tt x}, {\tt y}, and {\tt z} are respectively $\Number$,
$\Number$, and $\String$. This also means that the call to {\tt sum} compiles
without errors, as the first two components of the tuple are consistent with
the types of the parameters, and the other components are consistent with
$\Top$.

\subsection{Unions}

\subsection{Tables and Interfaces}

We need to explicitly annotate \texttt{next} being of type
\texttt{Element} $\cup\;\Nil$ because in our type system $\Nil$
is not the bottom type, that is, $\Nil$ is subtype of itself only.
Typed Lua has this behavior because in Lua $\Nil$ represents
the absence of an useful value.

\subsection{Modules}

\subsection{Objects and Classes}

\section{Related Work}
 \label{sec:review}

@@

\subsection{Similar Lua projects}

@@

Tidal Lock \citep{tidallock} is a prototype of another optional type
system for Lua, which is written in Metalua.
A remarkable feature of Tidal Lock is the way that it handles tables,
because it allows the programmer to incrementally build the type of
a table.
More precisely, Tidal Lock allows the programmer to create an empty
table to build its type according to the keys that the programmer
inserts in it, so the programmer does not need to build the table in
a single table constructor.
Typed Lua has a simpler form of this feature, with a different
formalization.
Like Typed Lua, Tidal Lock also has local type inference.
Unlike Typed Lua, Tidal Lock is not typing the common object-oriented
idioms used by Lua programmers.

\section{Conclusion} \label{sec:con}

>>>>

\newcommand{\Any}{\mathbf{any}}
\newcommand{\Nil}{\mathbf{nil}}
\newcommand{\False}{\mathbf{false}}
\newcommand{\True}{\mathbf{true}}
\newcommand{\Boolean}{\mathbf{boolean}}
\newcommand{\Number}{\mathbf{number}}
\newcommand{\String}{\mathbf{string}}
\newcommand{\Void}{\mathbf{void}}

@@

Dynamically typed languages such as Lua avoid static types in favor of
simplicity and flexibility, because the absence of static types means
that programmers do not need to bother about abstracting types that
should be validated by a type checker.
Instead, dynamically typed languages use run-time \textit{type tags}
to classify the values they compute, so their implementation can use
these tags to perform run-time (or dynamic) type checking
\citep{pierce2002tpl}.

This simplicity and flexibility allow programmers to write code that
might require a quite complex type system to statically type check,
though it may also hide bugs that will be caught only after deployment
if programmers do not properly test their code.
In contrast, static type checking helps programmers detect many 
bugs during the development phase.
Static types also provide a conceptual framework that helps
programmers define modules and interfaces that can be combined to
structure the development of programs.

Thus, early error detection and better program structure are two
advantages of static type checking that can lead programmers to
migrate their code from a dynamically typed to a statically
typed language, when their simple scripts become complex programs
\citep{tobin-hochstadt2006ims}.
Dynamically typed languages certainly help programmers during the
beginning of a project, because their simplicity and flexibility
allow quick development and make it easier to change code according to
changing requirements.
However, programmers tend to migrate from dynamically typed to
statically typed code as soon as the project has consolidated its
requirements, because the robustness of static types helps
programmers link requirements to abstractions.
This migration usually involves different languages that have
different syntax and semantics, which usually require a complete
rewrite of existing programs instead of incremental evolution from
dynamic to static types.

Ideally, programming languages should offer programmers the
option to choose between static and dynamic typing:
\textit{optional type systems} \citep{bracha2004pluggable} and
\textit{gradual typing} \citep{siek2006gradual} are two similar
approaches for blending static and dynamic typing in the same
language.
The aim of both approaches is to offer programmers the option
to use type annotations where static typing is needed and avoid them
where dynamic typing is sufficient.
The difference between these two approaches is the way they treat
run-time semantics.
While optional type systems do not affect the run-time semantics,
gradual typing uses run-time checks to ensure that dynamically typed
code does not violate the invariants of statically typed code.

Programmers and researchers sometimes use the term gradual typing
to mean the incremental evolution of dynamically typed code to
statically typed code.
For this reason, gradual typing may also refer to optional type
systems and other approaches that blend static and dynamic typing to
help programmers incrementally migrate from dynamic to static typing
without having to switch to a different language, though all these
approaches differ in the way they handle static and dynamic typing
together.
We use the term gradual typing for referring to the work of
\citet{siek2006gradual}.

In this work we propose the design and implementation of Typed Lua:
an optional type system for Lua that should be complex enough to
preserve some of the Lua idioms that programmers are already used to,
but that should also include new constructs that help programmers
structure Lua programs.

Lua is a small scripting language, so designing an optional type
system for Lua should shed some light on optional type systems for
scripting languages in general.
Lua provides mechanisms for organizing a program in modules and
writing object-oriented programs, but it does not set policy on how
these features should behave, due to its use as an embedded and
extension language.
Thus, implementing an optional type system for Lua should offer Lua
programmers one way to obtain most of the advantages of static typing
without compromising the simplicity and flexibility of dynamic typing.

So far, Typed Lua should be a strict superset of Lua that
provides optional type annotations, compile-time type checking, and
class-based object-oriented programming through the definition of
classes, interfaces, and modules.
In practice, we should implement Typed Lua as a programming
language that extends Lua syntax to add optional type annotations
and standard constructions for structuring Lua code.
The compiler should use static types to perform compile-time type
checking, but should also allow Lua code to coexist with Typed Lua
code, and generate Lua code that runs in unmodified Lua
implementations.
We have an ongoing implementation of the Typed Lua compiler that is
available online\footnote{https://github.com/andremm/typedlua}.

Typed Lua intended use is as an application language, and we view
that policies for organizing a program in modules and writing
object-oriented programs should be part of the language and
enforced by its optional type system.
An application language is a programming language that helps
programmers develop applications from scratch until these
applications evolve to complex systems rather than just scripts.

We also believe that Typed Lua will help programmers give a more
formal documentation to already existing Lua code, as static types
are also a useful source of documentation in languages that provide
type annotations, because type annotations are always validated by
the type checker and therefore never get outdated.
Thus, programmers can use Typed Lua to define axioms about the
interfaces and types of dynamically typed modules.
We will enforce this point by using Typed Lua to statically type
the interface of the Lua standard libraries and other commonly used
Lua libraries, so our compiler can check their usage by Typed Lua
code.

Typed Lua will perform a very limited form of local type inference
\citep{pierce2000lti}, as static typing does not necessarily mean
that programmers need to insert type annotations in the code.
Several statically typed languages such as Haskell provide some
amount of type inference that automatically deduces the types of
expressions.
Still, Typed Lua will need a small amount of type annotations due
to the nature of its optional type system.

Typed Lua will not deal with code optimization, although another
important advantage of static types is that they help the compiler
perform optimizations and generate more efficient code.
However, we believe that the formalization of our optional type
system will be precise enough to aid optimization in some Lua
implementations.

We will use some of the ideas of gradual typing to formalize Typed Lua.
Even though Typed Lua is an optional type system and thus will not
include run-time checks between dynamic and static regions of the
code, we believe that using the foundations of gradual typing to
formalize our optional type system will allow us to include run-time
checks in the future.

In Section \ref{sec:review} we review the literature about static and
dynamic typing in the same language, which includes the discussion
about the particularities between optional type systems and gradual
typing.
In Section \ref{sec:proposal} we describe the reasons why Typed Lua
includes features of object-oriented programming and the status of
its design and implementation.
In Section \ref{sec:con} we outline our expected contributions.

\section{Blending static and dynamic typing} \label{sec:review}

@@

We begin this section presenting statistics about the usage of Lua
features and idioms.
We collected statistics about how programmers use tables, functions,
dynamic type checking, object-oriented programming, and modules.
We shall see that these statistics informed important design decisions
on our optional type system.
We end this section introducing Typed Lua through examples, and
describing a few Lua projects that are similar to Typed Lua.

\subsection{Statistics about the usage of Lua}

We used the LuaRocks repository to build our statistics database;
LuaRocks \citep{hisham2013luarocks} is a package manager for Lua
modules.
We downloaded the 3928 \texttt{.lua} files that were available in
the LuaRocks repository at February 1st 2014.
However, we ignored files that are not compatible with Lua 5.2,
the latest version of Lua.
We also ignored \textit{machine-generated} files and test files,
because these files may not represent idiomatic Lua code,
and might skew our statistics towards non-typical uses of Lua.
This left 2598 \texttt{.lua} files from 262 different projects for
our statistics database;
we parsed these files and processed their abstract syntax tree
to gather the statistics that we show in this section.

To verify how programmers use tables, we measured how they
initialize, index, and iterate tables.
We present these statistics in the next three paragraphs to discuss
their influence on our type system.

The table constructor appears 23185 times.
In 36\% of the occurrences it is a constructor that initializes a
record (e.g., \texttt{\{ x = 120, y = 121 \}});
in 29\% of the occurrences it is a constructor that initializes a
list (e.g., \texttt{\{ "one", "two", "three", "four" \}});
in 8\% of the occurrences it is a constructor that initializes a
record with a list part;
and in less than 1\% of the occurrences (4 times) it is a constructor
that uses only the booleans \texttt{true} and \texttt{false} as indexes.
At all, in 73\% of the occurrences it is a constructor that uses
only literal keys;
in 26\% of the occurrences it is the empty constructor;
in 1\% of the occurrences it is a constructor with non-literal keys
only, that is, a constructor that uses variables and function calls
to create the indexes of a table;
and in less than 1\% of the occurrences (19 times) it is a constructor
that mixes literal keys and non-literal keys.

The indexing of tables appears 130448 times:
86\% of them are for reading a table field while
14\% of them are for writing into a table field.
We can classify the indexing operations that are reads as follows:
89\% of the reads use a literal string key,
4\% of the reads use a literal number key,
and less than 1\% of the reads (10 times) use a literal boolean key.
At all, 93\% of the reads use literals to index a table while
7\% of the reads use non-literal expressions to index a table.
It also worth mentioning that 45\% of the reads are actually
function calls.
More precisely, 25\% of the reads use literals to call a function,
20\% of the reads use literals to call a method, that is,
a function call that uses the colon syntactic sugar, 
and less than 1\% of the reads (195 times) use non-literal expressions
to call a function.
We can also classify the indexing operations that are writes as follows: 
69\% of the writes use a literal string key,
2\% of the writes use a literal number key,
and less than 1\% of the writes (1 time) uses a literal boolean key.
At all, 71\% of the writes use literals to index a table while
29\% of the writes use non-literal expressions to index a table.

We also measured how many files have code that iterate over tables to
observe how frequently iteration is used.
We observed that 23\% of the files have code that iterate over keys
of any value, that is, the call to \texttt{pairs} appears at least
once in these files (the median is twice per file);
21\% of the files have code that iterate over integer keys, that is,
the call to \texttt{ipairs} appears at least once in these files
(the median is also twice per file);
and 10\% of the files have code that use the numeric \texttt{for}
along with the length operator (the median is once per file).

The numbers about table initialization, indexing, and iteration
show us that tables are mostly used to represent records, lists,
and associative arrays.
Therefore, Typed Lua should include a table type for handling
these uses of Lua tables.
Even though the statistics show that programmers initialize tables
more often than they use the empty constructor to
dynamically initialize tables, the statistics of the empty
constructor are still expressive and indicate that Typed Lua should
also include a way to handle this style of defining table types.

We found a total of 24858 function declarations in our database
(the median is six per file).
Next, we discuss how frequently programmers use dynamic type
checking and multiple return values inside these functions.

We observed that 9\% of the functions perform dynamic type checking
on their input parameters, that is, these functions use \texttt{type}
to inspect the tags of Lua values (the median is once per function).
We randomly selected 20 functions to sample how programmers are
using \texttt{type}, and we got the following data:
50\% of these functions use \texttt{type} for asserting the tags of
their input parameters, that is, they raise an error when the tag of a
certain parameter does not match the expected tag, and
50\% of these functions use \texttt{type} for overloading, that is,
they execute different code according to the inspected tag.

These numbers show us that Typed Lua should include union types,
because the use of the \texttt{type} idiom shows that disjoint unions
will help programmers define data structures that can hold a value of
several different, but fixed types.
Typed Lua should also use \texttt{type} as a mechanism for decomposing
unions, though it may be restricted to base types only.

We observed that 10\% of the functions explicitly return multiple
values.
We also observed that 5\% of the functions return \texttt{nil} plus
something else, for signaling an unexpected behavior;
and 1\% of the functions return \texttt{false} plus something else,
also for signaling an unexpected behavior.

Typed Lua should include function types to represent Lua functions,
and tuple types to represent the signatures of Lua functions,
multiple return values, and multiple assignments.
Tuple types will require some special attention, because Typed Lua
should be able to adjust tuple types during compile-time, in a
similar way to what Lua does with function calls and multiple
assignments during run-time.
In addition, the number of functions that return \texttt{nil} and
\texttt{false} plus something else show us that overloading on the
return type will be useful to the type system.

We also measured how frequently programmers use the object-oriented
paradigm in Lua.
We observed that 23\% of the function declarations are actually
method declarations.
More precisely, 14\% of them use the colon syntactic sugar while
9\% of them use \texttt{self} as their first parameter.
We also observed that 63\% of the projects extend tables with
metatables, that is, they call \texttt{setmetatable} at least once,
and 27\% of the projects access the metatable of a given table,
that is, they call \texttt{getmetatable} at least once.
In fact, 45\% of the projects extend tables with metatables and
declare methods:
13\% using the colon syntactic sugar, 14\% using \texttt{self}, and
18\% using both.

Based on these observations, Typed Lua should include interfaces and
classes as syntactic sugar for table types that encode objects.
Proposing an optional type system for Lua that includes classes may
sound unusual at first glance, because Lua does not have standard
policies for object-oriented programming.
However, Lua provides mechanisms that allow programmers to abstract
their code in terms of objects, and our statistics confirm that
an expressive number of programmers are relying on these mechanisms to
use the object-oriented paradigm in Lua.
Typed Lua should include interfaces and classes as merely syntactic
mechanisms that help the compiler type check object-oriented code,
and they should not influence the semantics of Lua.

We also measured how programmers are defining modules.
We observed that 38\% of the files use the current way of defining
modules, that is, these files return a table that contains the
exported members of the module at the end of the file;
22\% of the files still use the deprecated way of defining modules,
that is, these files call the function \texttt{module};
and 1\% of the files use both ways.
At all, 61\% of the files are modules while 39\% of the files are
plain scripts.
The number of plain scripts is high considering the origin of
our database.
However, we did not ignore sample scripts, which usually serve to
help the users of a given module on how to use this module, and
that is the reason why we have a high number of plain scripts.

Based on these observations, Typed Lua should also use interfaces as
syntactic sugar for table types that define modules.
Typed Lua should also support the deprecated style of module
definition, using global names as exported members of the module.

Besides using interfaces for encoding objects and defining modules,
Typed Lua should also use interfaces for defining the types of
userdata, as programmers often use userdata in an object-oriented
style.

The last statistics that we collected were about variadic functions
and vararg expressions.
We observed that 8\% of the functions are variadic, that is,
their last parameter is the vararg expression.
We also observed that 5\% of the initialization of lists
(or 2\% of the occurrences of the table constructor) use solely the
vararg expression.
Typed Lua should include a \textit{vararg type} to handle variadic
functions and vararg expressions.

\subsection{Typed Lua through examples}

Our main goal is to allow programmers to combine static and dynamic
typing in Lua, so we propose Typed Lua as an extension to Lua 5.2
syntax that introduces optional type annotations, and class-based
object-oriented programming through the definition of classes,
interfaces, and modules.
An important requirement of Typed Lua is that it should be backwards
compatible with Lua, that is, it should allow programmers to run any
Lua code without modifications.
For this reason, these syntactic extensions include new keywords
that are not reserved words.

Figure \ref{fig:typelang} shows that Typed Lua includes first-class
and second-class types.
Typed Lua uses first-class types for representing Lua values and
second-class types for defining the signatures of functions.
First-class types include literal types, base types, the type $\Any$,
union types, function types, table types, and recursive types.
Second-class types include the type $\Void$, first-class types,
vararg types, and tuple types.
Literal types are types that represent a single value
(e.g., $\Nil$, $\False$, $\True$, literal numbers, and
literal strings).
Literal types are subtypes of their respective base types.
More precisely, the literals $\False$ and $\True$ are subtypes of
$\Boolean$, any literal number $n$ is subtype of $\Number$, and any
literal string $s$ is subtype of $\String$.
The type $\Any$ is the dynamic type that allows dynamic code coexist
with static code.
We will discuss union types, function types, table types, recursive
types, and second-class types along this section.

\begin{figure}[!ht]
\textbf{Type Language}\\
\dstart
$$
\begin{array}{rlr}
T ::= & & \textit{first-class types}\\
& \;\; L & \textit{literal types}\\
& | \; B & \textit{base types}\\
& | \; \Any & \textit{dynamic type}\\
& | \; T \cup T & \textit{union type}\\
& | \; S \rightarrow S & \textit{function type}\\
& | \; \{T:T, ..., T:T\} & \textit{table type}\\
& | \; X & \textit{type variable}\\
& | \; \mu X.T & \textit{recursive type}\\
L ::= & \;\; \Nil \; | \; \False \; | \; \True \\
& | \; {<}{\it number}{>} \; | \; {<}{\it string}{>} & \\
B ::= & \Boolean \; | \; \Number \; | \; \String & \\
S ::= & \Void \; | \; V & \textit{second-class types}\\
V ::= & \;\; T & \\
& | \; T* & \textit{vararg type}\\
& | \; T \times V & \textit{tuple type}
\end{array}
$$
\dend
\caption{Typed Lua types}
\label{fig:typelang}
\end{figure}

Typed Lua allows optional type annotations in variable and function
declarations.
Whenever it is possible to use local type inference, the compiler
assigns a more specific type to an unannotated declaration.
When it is not possible to use local type inference, the compiler
assigns the type $\Any$ to an unannotated declaration.
For instance, in function declarations, the compiler does not infer
a more specific type to unannotated input parameters, because,
in this case, global type inference often is undecidable in a
language with subtyping and overloading \citep{wells1999typability}.

In the following example we use type annotations in a function
declaration, and we do not use type annotations in a declaration of a
local variable:
\begin{verbatim}

    local function fact (n:number):number
      if n == 0 then
        return 1
      else
        return n * fact(n - 1)
      end
    end
    local x = 5
    print(fact(x))
\end{verbatim}

Even though we did not add type annotations in the declaration of
the local variable \texttt{x}, the compiler uses local type inference to
assign the type $\Number$ to \texttt{x}.
The compiler does not use the literal type $5$ instead of the base
type $\Number$ because this would prevent
programmers from assigning other numeric values to the variable
\texttt{x}.
The inference that we implement in Typed Lua is quite simple, as it
uses only the type of the local expression.

Typed Lua allows programmers to combine annotated code with
unannotated code, as we show in the following example:
\begin{verbatim}
    local function abs (n:number):number
      if n < 0 then
        return -n
      else
        return n
      end
    end

    local function dist (x, y)
      return abs(x - y)
    end
\end{verbatim}

The function \texttt{dist} receives two parameters of type $\Any$
and returns a value of type $\Number$.
The compiler assigns the dynamic type $\Any$ to the input
parameters of \texttt{dist} because they do not have type annotations,
and the compiler does not use global type inference, as we mentioned
previously.
The compiler infers the return type of \texttt{dist} because it is
not a recursive function, that is, Typed Lua does not infer the return
type of recursive functions.

We are using the consistent-subtyping relation of gradual
typing to check the interaction among the dynamic type and the other
types.
More precisely, we can call \texttt{abs} inside \texttt{dist} because
\texttt{x - y} is of type $\Any$, and the type $\Any$ is
consistent with the type $\Number$.
The type $\Any$ is not the union of all types, but it is a separate
type.

Unlike gradual typing, we do not insert run-time checks that
inspect the interaction between dynamically typed code and statically
typed code during run-time.
In other words, Typed Lua does not guarantee that dynamically typed
code does not violate statically typed code.
Typed Lua does not insert run-time checks because they can decrease
run-time performance \citep{allende2013cis}.
We believe that a careful evaluation of run-time checks should be done
before inserting them on the type system.
In the previous example, we cannot guarantee that \texttt{dist} is never
going to call \texttt{abs} with a parameter that is not a number,
because in the semantics of Lua the minus operator can result in a
value that is not a number.
Still, we can implement run-time checks in the future because we are
using the consistent-subtyping relation of gradual typing to
formalize our optional type system, and the type system is being
designed to be sound.

We need to explicitly annotate \texttt{next} being of type
\texttt{Element} $\;\cup\;\Nil$ because in our type system $\Nil$
is not the bottom type, that is, $\Nil$ is subtype of itself only.
Typed Lua has this behavior because in Lua $\Nil$ represents
the absence of an useful value.

Typed Lua includes function types because functions are first-class
values in Lua.
The syntax for defining a function type is \texttt{(s) -> (s)},
where \texttt{s} is a second-class type.
As we saw in Figure \ref{fig:typelang}, a second-class type in
Typed Lua can be the type $\Void$, a first-class type, a variadic
type, or a tuple type.
We use tuple types to represent multiple parameters, multiple return
values, and multiple assignments, and we use the variadic type to
represent variadic functions and the vararg expression.
Even though we already included the tuple type and the vararg type,
we are still working on these types.

The next example shows how we can annotate a \texttt{map} function
that receives another function as its parameter and uses this
function to alter the elements of a list of numbers, which is
also a parameter of \texttt{map}:
\begin{verbatim}
    local function map (f:(number)->(number),
                        l:{number}):void
      for i=1, #l do
        l[i] = f(l[i] or 0)
      end
    end
\end{verbatim}

To conclude this section, we just started the design and
implementation of Typed Lua, and there is still several design
decisions to be taken on this subject.
For instance, our type system should handle the subtleties that
come with metatables.

\subsection{Related Lua projects}

@@

Tidal Lock \citep{tidallock} is a prototype of another optional type
system for Lua, which is written in Metalua.
A remarkable feature of Tidal Lock is the way that it handles tables,
because it allows the programmer to incrementally build the type of
a table.
More precisely, Tidal Lock allows the programmer to create an empty
table to build its type according to the keys that the programmer
inserts in it, so the programmer does not need to build the table in
a single table constructor.
Typed Lua has a simpler form of this feature, with a different
formalization.
Like Typed Lua, Tidal Lock also has local type inference.
Unlike Typed Lua, Tidal Lock does not type the common object-oriented
idioms used by Lua programmers.

\section{Expected contributions} \label{sec:con}

Our main contribution is the design of a complete optional type system
for Lua, a procedural scripting language.
TypeScript is a complete optional type system for JavaScript,
but it does not include union types, and it uses arrays to represent
variadic functions and multiple return values. 
Gradualtalk is a complete gradual type system for Smalltalk,
but Smalltalk is an object-oriented scripting language that does
not handle multiple return values, multiple assignments, and
variadic functions.
Gradualtalk also mixes nominal and structural typing because of
the object-oriented nature of Smalltalk.
Typed Racket is a complete static type system for Racket, a functional
scripting language.
Typed Racket makes it easier to migrate from untyped to typed code,
but the migration happens module-by-module and type annotations are
mandatory on typed modules.
There are research on the design of optional and gradual type systems
for JavaScript, Perl, and Python, but they are either experimental or
limited in their coverage.
Furthermore, Typed Lua includes table types that use types as keys,
instead of identifiers, and the fact that Typed Lua also includes
literal types makes its table types subsume the usual idea of
record types.

Although each language is different, we believe that several parts of
the design of Typed Lua should be relevant to other scripting languages.
Most notably, JavaScript shares with Lua the use of tests of type tags
as a way of encoding optional and overloaded parameters, and the
proliferation of class-based object systems built on top of the
language's delegation mechanisms.

Typed Lua should also be a major contribution to the Lua community,
because it offers a framework that programmers can use to document,
test, and better structure their applications.
For libraries where a full conversion to static type checking should
prove unfeasible or too much work, the community can use Typed Lua
just to document the external interfaces of the libraries,
giving the benefits of static type checking to the users of these
libraries.

Finally, Typed Lua can be a base for further research and development
of static type systems for the Lua language, such as the use of static
analysis of Lua API calls for checking that a native library is
exporting the interface to Lua code that it declares, the use of
effect types for typing Lua coroutines, and the use of static types
for code optimization.
